Bayesian spectral density estimation using P-splines with quantile-based knot placement

### 2020-07-05 pSplines.tex
starting value w text

This article  proposes a Bayesian approach to estimating the spectral density of a stationary time series using a prior based on a mixture of P-spline distributions.  Our proposal is motivated by the B-spline Dirichlet process prior of \cite{Edwards2019} in combination  with Whittle's likelihood and aims at  reducing the high computational complexity of its posterior computations. The strength of the B-spline Dirichlet process prior over the Bernstein-Dirichlet process prior of \cite{Choudhuri:2004}  lies in its ability to estimate spectral densities with sharp peaks and abrupt changes due to the flexibility of B-splines with variable number and location of knots.Here, we suggest to use P-splines of \cite{Eilers:1996} that combine a B-spline basis with a discrete penalty on the basis coefficients.  \textcolor{red}{In addition to equidistant knots, a novel strategy for a more expedient placement of  knots is proposed that makes use of the information provided by the periodogram about the steepness of the spectral power distribution.}
We demonstrate in a simulation study and two real  case studies that this approach retains the flexibility of the B-splines, achieves similar ability to accurately estimate peaks due to the new data-driven knot allocation scheme but significantly reduces the computational costs.

@@

Bayesian nonparametric prior specifications for the psd include \cite{Cart:1997}
who used a smoothing prior on the log psd.  \cite{Rosen:2012} segmented a nonstationary time series into stationary segments. \textcolor{red}{For each stationary segment, they decomposed the log psd into a linear and nonlinear component and  put a linear smoothig spline prior on the nonlinear part. }\cite{Pensky:2007} propose Bayesian wavelet-smoothing of the log psd and recently, \cite{Cadonna2017} modelled the log psd  as a mixture of Gaussian distributions with frequency-dependent weights and mean functions.
%
\cite{Gangopadhyay:1999} modelled the log-spectral density function by  piecewise polynomials of low order between strategically placed knots on the support of the psd,  allowing the data to select the number and location of the knots.  A reversible jump Markov chain Monte Carlo \citep[RJMCMC;][]{Green:1995} algorithm was implemented to sample from the posterior distribution.  Similarly, frequentist approaches such as \cite{Rodriguez} and \cite{Wood2017} model the log spectral density by a linear combination of splines with variable knots and combine these with adaptive smoothing.

Penalized splines have been used for a time-domain analysis of time series data e.g.\ to model the mean as a smooth function of time by \cite{Krivobokova}  or for fitting and  forecasting univariate  nonlinear time series by \cite{Wegener}, but not in the frequency domain for estimating the spectral density.
The P-spline prior on the spectral density function that we are going to propose is related to the 
Bernstein-Dirichlet process (BDP) prior  introduced by \cite{Choudhuri:2004} and combined with the Whittle likelihood.  The BDP prior  is a mixture of beta densities with weights induced by a Dirichlet process.  The number of components, which is a smoothing parameter, is given a discrete prior.  \cite{Edwards2019} extended the method by replacing the beta densities with B-spline densities. Similar to 
\cite{Gangopadhyay:1999}, the specification of the number and location of the B-spline densities is data-driven, however it avoids a RJMCMC algorithm by putting a second Dirichlet process prior  on the cdf that induces the inter-knot distances. \textcolor{red}{ \cite{Choudhuri:2004} already demonstrated in a comprehensive simulation study  that the Bayesian psd estimates based on the BDP prior outperforms in terms of the $L_1$-error the smoothed periodogram estimates based on Bartlett-Priestley quadratic kernel, the penalized MLE of \cite{Pawitan} and autoregressive spectral estimates in most scenarios but noted that the BDP-based estimates detect peaks correctly but often underestimate the magnitude of sharp peaks.}
Unlike the beta densities  which have global support on the interval [0,1], B-spline densities have local support, thereby increasing their flexibility and allowing better modelling of spectral densities with sharp peaks and abrupt changes.  As demonstrated in \cite{Edwards2019}, the B-spline--Dirichlet process (BspDP) prior outperforms the BDP prior in estimating complex psd's.  However, this flexibility comes at a high computational price.

@@

$\bm{\xi}$ contains fixed equidistant internal knots on $[0,1]$ in our basic P-spline prior. \textcolor{red}{A more judicious scheme for the placement  of knots that not only speeds up posterior computations but improves the fit for peaked spectral densities is described in  section ``Quantile-based knot placement".}
%Therefore, the P-spline prior on the spectral density function is given by
%\begin{align*}
%f(\pi \omega) = \tau \times s_r(\omega;\bm{w}, \bm{\xi}), \quad \omega \in [0,1],
%\end{align*}
%where $\tau = \int_{0}^{1}f(\tau \omega)\text{d}\omega$ is the normalizing constant.

@@

However, to sample from the full conditional distribution of $\bm{v}$, we use the Metropolis algorithm.
In order to improve the mixing of the chains, we apply the Metropolis algorithm on a reparametrized posterior distribution \citep{Lambert:2007}.  For this, we need a pilot posterior sample for $\bm{v}$ from which we calculate its mean vector $\bm{\overline{v}}$ and covariance matrix $\textbf{S}$.  Then we define the following re-parametrization
\begin{align*}
\bm{v} = \textbf{S}^{1/2} \bm{\beta} + \overline{\bm{v}},
\end{align*}
where $\bm{\beta}=(\beta_1, \dots, \beta_{K-1})^\top$.

@@

The knots are often equally spaced in P-spline methods.  This scheme works quite well  for smooth psds without any abrupt changes.  However, for spectral densities with spikes, the algorithm will require a large number of knots to adequately estimate the peaks.  The B-spline prior method is able to handle abrupt changes because the knot location is variable and  driven by the data.  We propose a new selective method that allocates the knots according to quantiles of a periodogram-related distribution function but once allocated, the knots stay fixed.  The idea is to concentrate more effort in those regions that have potential peaks as detected by the periodogram.  \textcolor{red}{It is similar in spirit to the knot placement of O-splines \cite{WandOrmerod} in the regression context that make use of the empirical quantiles of the covariates. }

>>>>

This article  proposes a Bayesian approach to estimating the spectral density of a stationary time series using a prior based on a mixture of P-spline distributions.  Our proposal is motivated by the B-spline Dirichlet process prior of \cite{Edwards2019} in combination  with Whittle's likelihood and aims at  reducing the high computational complexity of its posterior computations. The strength of the B-spline Dirichlet process prior over the Bernstein-Dirichlet process prior of \cite{Choudhuri:2004}  lies in its ability to estimate spectral densities with sharp peaks and abrupt changes due to the flexibility of B-splines with variable number and location of knots. Here, we suggest to use P-splines of \cite{Eilers:1996} that combine a B-spline basis with a discrete penalty on the basis coefficients.  In addition to equidistant knots, a novel strategy for a more expedient placement of  knots is proposed that makes use of the information provided by the periodogram about the steepness of the spectral power distribution.
We demonstrate in a simulation study and two real  case studies that this approach retains the flexibility of the B-splines, achieves similar ability to accurately estimate peaks due to the new data-driven knot allocation scheme but significantly reduces the computational costs.

@@

Bayesian nonparametric prior specifications for the psd include \cite{Cart:1997}
who used a smoothing prior on the log psd.  \cite{Rosen:2012} segmented a nonstationary time series into stationary segments.  For each stationary segment, they decomposed the log psd into a linear and nonlinear component and  put a linear smoothig spline prior on the nonlinear part.  \cite{Pensky:2007} propose Bayesian wavelet-smoothing of the log psd and recently, \cite{Cadonna2017} modelled the log psd  as a mixture of Gaussian distributions with frequency-dependent weights and mean functions.
%
\cite{Gangopadhyay:1999} modelled the log-spectral density function by  piecewise polynomials of low order between strategically placed knots on the support of the psd,  allowing the data to select the number and location of the knots.  A reversible jump Markov chain Monte Carlo \citep[RJMCMC;][]{Green:1995} algorithm was implemented to sample from the posterior distribution.  Similarly, frequentist approaches such as \cite{Rodriguez} and \cite{Wood2017} model the log spectral density by a linear combination of splines with variable knots and combine these with adaptive smoothing.

Penalized splines have been used for a time-domain analysis of time series data e.g.\ to model the mean as a smooth function of time by \cite{Krivobokova}  or for fitting and  forecasting univariate  nonlinear time series by \cite{Wegener}, but not in the frequency domain for estimating the spectral density.
The P-spline prior on the spectral density function that we are going to propose is related to the 
Bernstein-Dirichlet process (BDP) prior  introduced by \cite{Choudhuri:2004} and combined with the Whittle likelihood.  The BDP prior  is a mixture of beta densities with weights induced by a Dirichlet process.  The number of components, which is a smoothing parameter, is given a discrete prior.  \cite{Edwards2019} extended the method by replacing the beta densities with B-spline densities. Similar to 
\cite{Gangopadhyay:1999}, the specification of the number and location of the B-spline densities is data-driven, however it avoids a RJMCMC algorithm by putting a second Dirichlet process prior  on the cdf that induces the inter-knot distances. \cite{Choudhuri:2004} already demonstrated in a comprehensive simulation study  that the Bayesian psd estimates based on the BDP prior outperforms in terms of the $L_1$-error the smoothed periodogram estimates based on Bartlett-Priestley quadratic kernel, the penalized MLE of \cite{Pawitan} and autoregressive spectral estimates in most scenarios but noted that the BDP-based estimates detect peaks correctly but often underestimate the magnitude of sharp peaks.
Unlike the beta densities  which have global support on the interval [0,1], B-spline densities have local support, thereby increasing their flexibility and allowing better modelling of spectral densities with sharp peaks and abrupt changes.  As demonstrated in \cite{Edwards2019}, the B-spline--Dirichlet process (BspDP) prior outperforms the BDP prior in estimating complex psd's.  However, this flexibility comes at a high computational price.

@@

$\bm{\xi}$ contains fixed equidistant internal knots on $[0,1]$ in our basic P-spline prior.  A more judicious scheme for the placement  of knots that improves the fit for peaked spectral densities is described in  section ``Quantile-based knot placement".
%Therefore, the P-spline prior on the spectral density function is given by
%\begin{align*}
%f(\pi \omega) = \tau \times s_r(\omega;\bm{w}, \bm{\xi}), \quad \omega \in [0,1],
%\end{align*}
%where $\tau = \int_{0}^{1}f(\tau \omega)\text{d}\omega$ is the normalizing constant.

@@

However, to sample from the full conditional distribution of $\bm{v}$, we use the Metropolis algorithm.
\textcolor{red}{Since the weights in general are larger in those areas of the frequency domain where the psd has peaks, we specify their starting value proportionally to the periodogram. Applying the corresponding transformation we obtain the starting value for the vector $\bm{v}$.  This strategy speeds up the MCMC process by shorting the burn-in period.}

In order to improve the mixing of the chains, we apply the Metropolis algorithm on a reparametrized posterior distribution \citep{Lambert:2007}.  For this, we need a pilot posterior sample for $\bm{v}$ from which we calculate its mean vector $\bm{\overline{v}}$ and covariance matrix $\textbf{S}$.  Then we define the following re-parametrization
\begin{align*}
\bm{v} = \textbf{S}^{1/2} \bm{\beta} + \overline{\bm{v}},
\end{align*}
where $\bm{\beta}=(\beta_1, \dots, \beta_{K-1})^\top$.

@@

The knots are often equally spaced in P-spline methods.  This scheme works quite well  for smooth psds without any abrupt changes.  However, for spectral densities with spikes, the algorithm will require a large number of knots to adequately estimate the peaks.  The B-spline prior method is able to handle abrupt changes because the knot location is variable and  driven by the data.  We propose a new selective method that allocates the knots according to quantiles of a periodogram-related distribution function but once allocated, the knots stay fixed.  The idea is to concentrate more effort in those regions that have potential peaks as detected by the periodogram.  It is similar in spirit to the knot placement of O-splines \citep{WandOrmerod} in the regression context that make use of the empirical quantiles of the covariates.
